Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-12 dev/paper/sections/discussion.qmd
some work on rephrasing narrative in intro
almost done, findings still need to be fleshed out though

# Experiments {#sec-empirical}

1. Shift of focus from individual to group of individuals (related: https://www.researchgate.net/publication/353073138_Generating_Collective_Counterfactual_Explanations_in_Score-Based_Classification_via_Mathematical_Optimization)
2. Convergence criterium matters: terminating once threshold probability is reached may not be optimal (see e.g. REVISE)
3. Optimizer choice matters: dimensionality is typically low, so no obvious benefit to using ADAM.
    - This might be better placed in JuliaCon proceedings, perhaps backed by small blog post on the matter. 
4. Mitigating strategy: penaliye distance from centroid.

>>>>

# Discussion {#sec-discussion}

Our results in @sec-empirical-2 indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice. These induced shifts can and should be considered as an (expected) external cost of individual recourse. While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making. We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments. In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool. As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether. This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

So, where does this leave us? We would argue that the expected external costs of individual recourse should be shared by all stakeholders. The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in @eq-collective. This will on average lead to more costly counterfactual outcomes. But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals. We have shown various types of shift-mitigating strategies that can be used to this end. Since all of these strategies can be seen simply as specific adaption of @eq-collective, they can be applied to any of the various counterfactual generators studied here.

### 2022-08-12 dev/paper/sections/empirical_2.qmd
almost done, findings still need to be fleshed out though

@@

Our findings ...

>>>>

In light of the huge number of experiments we have conducted, we shall start this section off with a few high-level observations. Across all datasets (synthetic and real), classifiers and counterfactual generators we observe either most or all of the following dynamics at varying degrees:

- Statistically significant domain and model shift as measured by MMD. 
- A deterioration in out-of-sample model performance as measured by the F-Score evaluated on a test sample. In many cases this drop in performance is substantial.
- Significant perturbations to the model parameters as well as an increase in the model's decisiveness.
- Disagreement between the original and retrained model, in some cases large.

There is also some clear heterogeneity across the results:

- The observed, adverse dynamics are generally speaking of the highest magnitude for the simple linear classifier. Differences in results for the MLP and Deep Ensemble are mostly negligible. 
- The reduction in model performance appears to be most severe when classes are not perfectly separable.
- With the exception of the Greedy generator, all other generators generally perform somewhat better than Wachter overall as expected.

Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix. Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above. @fig-syn presents a small subset of our results for the synthetic dataset with overlapping classes. It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds. The top row shows the estimated domain shifts. While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds. With respect to the domain shifts, the Greedy generator actually induces the smallest shifts. In general, we have observed the opposite. The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier. As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small. In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy. The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.

![Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.](www/synthetic_results.png){#fig-syn fig.pos="h" width="9cm" height="9cm"}

Turning to the real-world data next ...

![PLACEHOLDER: Results for real-world as table.](www/placeholder.png){#fig-real fig.pos="h" width="9cm" height="9cm"}

@@

Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts (@fig-mitigate-results) ...

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR.

![PLACEHOLDER: Results for mitigating strategies.](www/synthetic_results.png){#fig-mitigate-results fig.pos="h" width="9cm" height="9cm"}

### 2022-08-12 dev/paper/sections/introduction.qmd
almost done, findings still need to be fleshed out though

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of principal-agent problem in the context of data-driven decision-making. It involves a group of **principals** - i.e. human stakeholders - that fail to understand the behaviour of their **agent** - i.e. the model used for automated decision-making [@borch2022machine].

Models or algorithms that fall into this category are typically referred to **black-box** models. Despite their shortcomings, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons]. The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.

@@

![Dynamics in Algorithmic Recourse: we have a simple Bayesian model trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a domain shift (b); as the classifier is retrained we observe a model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}

@@

The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature. @sec-method posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs. @sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR. @sec-empirical presents our experimental results for synthetic and real-world datasets. We also present evidence for the effectiveness of various proposed mitigation strategies in that section. Our findings are then discussed in the broader context of the literature in @sec-discussion. We also point to some of the limitations of our work as well as avenues for future research in @sec-limit. Finally, @sec-conclusion concludes.

>>>>

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making. They are essentially **black boxes**: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools. Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes [@o2016weapons]. The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models [@rudin2019stop], others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as **interpretable AI** and **explainable AI** (XAI), respectively.

@@

![Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}

@@

The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature. @sec-method posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs. @sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR. @sec-empirical presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed. @sec-empirical-2 presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies. Our findings are then discussed in the broader context of the literature in @sec-discussion, before pointing to some of the limitations of our work as well as avenues for future research in @sec-limit. Finally, @sec-conclusion concludes.

### 2022-08-12 dev/paper/sections/limitations.qmd
almost done, findings still need to be fleshed out though

While we believe that this work constitutes a valuable starting point for addressing existing issues in Algorithmic Recourse from a fresh perspective, we are aware of several of its limitations. In the following we highlight some of these limitations and point to avenues for future research.

>>>>

While we believe that this work constitutes a valuable starting point for addressing existing issues in Algorithmic Recourse from a fresh perspective, we are aware of several of its limitations. In the following, we highlight some of these limitations and point to avenues for future research.

## Private vs. External Costs

Perhaps the most crucial shortcoming of our work is that we merely point out that there exists a trade-off between private costs to the individual and external costs to the collective of stakeholders. We fall short of providing any definite answers as to how that trade-off may be resolved in practice. The mitigation strategies we have proposed here provide a good starting point, but they are ad-hoc, mechanical extensions of the existing AR framework. An interesting idea to explore in future work could be the potential for Pareto optimal Algorithmic Recourse, that is a collective recourse outcome in which no single individual can be made better off, without make at least one other individual worse off. This type of work would be interdisciplinary and could help to formalize some of the concepts presented in this work. For the time being, we suggest that the trade-off can be resolved on a case-by-case basis.

### 2022-08-12 dev/paper/sections/methodology_2.qmd
almost done, findings still need to be fleshed out though

Finally, we introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample. We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function. This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes. Although the exact values for this metric are not important for our study, they can be used to detect model shifts. If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.

>>>>

We further introduce **Decisiveness** as a metric that quantifies the likelihood that a model assigns a high probability to its classification of any given sample. We define the metric simply as ${1\over{N}}\sum_{i=0}^N(\sigma(M(x)) − 0.5)^2$ where $M(x)$ are predicted logits from a binary classifier and $\sigma$ denotes the sigmoid function. This metric provides an unbiased estimate of the binary classifier's tendency to produce high-confidence predictions in either one of the two classes. Although the exact values for this metric are not important for our study, they can be used to detect model shifts. If decisiveness changes over time, then this is indicative of the decision boundary moves towards either one of the two classes.

Finally, we also take a look at the out-of-sample **Performance** of our models. To this end, we compute their F-score on a test sample that we leave untouched throughout the experiment.
